---
title: "Penguins"
output: html_document
date: "2025-09-26"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries

Here are the libraries used for this viusualization and analysis.

```{r}
library(datasets)
library(ggplot2)
library(tidyverse)
library(dplyr)
library(tibble)
library(palmerpenguins)
library(BSDA)
library(ggcorrplot)

```
## Our Dataset

This dataset contains 344 observations with 7 factors:

Species: Denotes penguin species
Island: Denotes an island in the Palmer Archipelago where the penguin was 
observed.
Bill_length_mm:  Bill length in millimeters
Bill_depth_mm: Bill Depth in millimeters
Flipper_length_mm: Flipper Length in millimeters
Body_mass_g: Body mass in grams
Sex: Denotes penguin sex (Male/Female)
Year: Denotes study year(2007,2008, or 2009)

```{r}
count(penguins)
summary(penguins)
head(penguins)
```




## Including Plots


```{r}
penguins %>% 
  ggplot(aes(x = species, fill = sex)) + 
  geom_bar(position = "dodge") +
  theme_bw()
  
  
```
Suppose we want to analyze if there analyze the factors behind penguin mass.
For example, let's see if flipper length is a significant factor in predicting
body mass. In this graph, We can see that there seems to be some positive 
relationship between flipper length and body mass. Adelie and Chinstrap Penguins
seem to have similar body mass and dimensions, and are on average smaller
than Gentoo penguins.
```{r pressure, echo=FALSE}
penguins %>%
  ggplot(aes(x = flipper_length_mm, y = body_mass_g, 
                        color = species, shape = species)) +
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) +
  xlab("Flipper Length (mm)") +
  ylab("Body Mass (g)") +
  ggtitle("Flipper Length vs Body Mass Across Penguin Species") +
  labs(color = "Species", shape = "Species") +
  theme_bw()
penguins %>%
  ggplot(aes(x = bill_length_mm, y = body_mass_g, 
                        color = species, shape = species)) +
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) +
  xlab("Bill Length (mm)") +
  ylab("Body Mass (g)") +
  ggtitle("Bill Length vs Body Mass Across Penguin Species") +
  labs(color = "Species", shape = "Species") +
  theme_bw()
penguins %>%
  ggplot(aes(x = bill_depth_mm, y = body_mass_g, 
                        color = species, shape = species)) +
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) +
  xlab("Bill Depth (mm)") +
  ylab("Body Mass (g)") +
  ggtitle("Bill Depth vs Body Mass Across Penguin Species") +
  labs(color = "Species", shape = "Species") +
  theme_bw()



```

If we do a simple regression analysis, we can see that flipper length is a
significant predictor for body mass, with an extremely small p-value well
below our significance level of 0.05. With ever 1mm increase with flipper length,
we can expect body mass to increase by 49 grams
```{r}
penguin.body.mass <- lm(body_mass_g ~ flipper_length_mm, data = penguins)
summary(penguin.body.mass)
```
Suppose we want to do a multilinear model using all continous variables: 
Flipper length, bill length, and bill depth. If we want to do a multilinear model, 
we should analyze the normality, homoscedacity, and multicollinearity between 
these variables first. We can see that there is a small correlation betwen bill 
length and flipper length, but not large enough to remove either of them based on
multicollinearity alone.

```{r}
penguin.body.mass.full <- lm(body_mass_g ~ flipper_length_mm + bill_length_mm+bill_depth_mm, data = penguins)
penguins.clean <- na.omit(penguins)
penguins.reduced <- penguins.clean %>% subset(select = c(flipper_length_mm, bill_length_mm, bill_depth_mm))
penguins.reduced
penguins_corr <- cor(penguins.reduced)
ggcorrplot(penguins_corr, hc.order = TRUE, type = "lower", lab = TRUE)
```

Let's also take a look at our residuals.  Looking at our qq plot, we can see
that our residuals line out nicely, suggesting that they follow a normal districution.

```{r}
qqnorm(penguin.body.mass.full$residuals)
qqline(penguin.body.mass.full$residuals)
```

We have fulfilled all the assumptions for using multicollinearity, but is the
new model actually more effective than the old one? Using ANOVA and a P-value of
0.05, we reach a p-value of 0.17, which means we fail to reject the null. We cannot
conclude that the full model is a better fit than the reduced model.
```{r}
?anova
anova(penguin.body.mass.full,penguin.body.mass)
```





