---
title: "Penguins"
output:
  pdf_document: default
  html_document: default
date: "2025-09-26"
---

```{r setup, include=FALSE}
#tinytex::install_tinytex()
knitr::opts_chunk$set(echo = FALSE)
```
## Introduction

This is a simple analysis of the Penguins dataset from the Palmer Penguins dataset.
This analysis will go over linear regression, data visualization, anova, and t-testing. 
If you would like at the code for this project, please visit the markdown file.



## Libraries

Here are the libraries used for this viusualization and analysis.

```{r echo=TRUE, results='hide'}
library(datasets)
library(ggplot2)
library(tidyverse)
library(dplyr)
library(tibble)
library(palmerpenguins)
library(BSDA)
library(ggcorrplot)
```
## Our Dataset

This dataset contains 344 observations with 7 factors:  

Species: Denotes penguin species  
Island: Denotes an island in the Palmer Archipelago where the penguin was
observed.  
Bill_length_mm:  Bill length in millimeters.  
Bill_depth_mm: Bill Depth in millimeters.  
Flipper_length_mm: Flipper Length in millimeters.  
Body_mass_g: Body mass in grams.  
Sex: Denotes penguin sex (Male/Female).  
Year: Denotes study year(2007,2008, or 2009)  

```{r}
summary(penguins)
head(penguins)
```




## Initial Visualizations

Before we dive into some regression analysis, let's get an overall understanding
of what our data looks like and how it is distributed.  With these graphs, we can
see that we have a roughly equal amount of male and female penguins. Adelie penguins
are the most plentiful, and are also the only species found on every island.

```{r}
penguins.clean <- na.omit(penguins)
penguins.clean %>% 
  ggplot(aes(x = species, fill = sex)) + 
  geom_bar(position = "dodge") +
  xlab("Island") +
  ylab("Amount") +
  labs(fill = "Sex") +
  theme_bw()

penguins.clean %>% 
  ggplot(aes(x = island, fill = species)) + 
  geom_bar(position = "dodge") +
  xlab("Island") +
  ylab("Amount") +
  labs(fill = "Species") +
  theme_bw()
  
  
```


Suppose we want to analyze if there analyze the factors behind penguin mass.
The following are graphs depicting the relationship between body dimensions and
body mass across all species of penguins. Adelie and Chinstrap Penguins
seem to have similar dimensions in regards to flipper length and bill depth.
We can also see that there seems to be some positive 
relationship between body dimenstions and body mass. 


```{r pressure, echo=FALSE}
penguins.clean %>%
  ggplot(aes(x = flipper_length_mm, y = body_mass_g, 
                        color = species, shape = species)) +
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) +
  labs(color = "Species", shape = "Species", x = "Flipper Length (mm)", 
       y = "Body Mass (g)", title = "Flipper Length vs Body Mass Across Penguin Species") +
  theme_bw()

penguins.clean %>%
  ggplot(aes(x = bill_length_mm, y = body_mass_g, 
                        color = species, shape = species)) +
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) +
   labs(color = "Species", shape = "Species", x = "Bill Length (mm)", 
       y = "Body Mass (g)", title = "Bill Length vs Body Mass Across Penguin Species") +
  theme_bw()

penguins.clean %>%
  ggplot(aes(x = bill_depth_mm, y = body_mass_g, 
                        color = species, shape = species)) +
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) +
  labs(color = "Species", shape = "Species", x = "Bill Depth (mm)", 
       y = "Body Mass (g)", title = "Bill Depth vs Body Mass Across Penguin Species") +
  labs(color = "Species", shape = "Species") +
  theme_bw()



```
## Model Creation and Analysis

Now let's create some models that can help us predict certian dimensions or 
features of a given penguin. For these models, let us assume asignificance level of 0.05. 
Suppose we wanted to predict a penguin's body mass using its other body dimensions.
Since different species will have different dimensions, we should focus on one species at a time.
In the above visualizations, we can observe that the Gentoo are isolated from the other species regarding body mass, 
let's focus on them for now. Let's start with a simple regression analysis
using flipper length. If we do a simple regression analysis, we can see 
that flipper length is a significant predictor for body mass, with an extremely 
small p-value well below our significance level of 0.05. We can reject the null and
assume the alternative is true: flipper length is a significant predictor for 
body mass. Our model is as follows:  

$$BodyMass = -6674.20 + 54.12FlipperLength$$


With ever 1mm increase  with flipper length for a Gentoo penguin, 
we can expect body mass to increase by 54 grams.
```{r}
gentoo.penguins <- penguins.clean %>% filter(species == "Gentoo")
gentoo.body.mass <- lm(body_mass_g ~ flipper_length_mm, data = gentoo.penguins)
summary(gentoo.body.mass)
```
Suppose we wanted to include more factors into this model and make it a multilinear model, 
such as Bill Length and Bill Depth. In order to do this, we have to assume that Bill Depth and Bill Length 
do not have too much correlation with the other variables, and also have a linear relationship with Body Mass,
Our scatterplots above demonstrate the latter, so we should verify the former through a correlation matrix

```{r}
penguins.reduced <- penguins.clean %>% subset(select = c(flipper_length_mm, bill_length_mm, bill_depth_mm))
penguins.corr <- cor(penguins.reduced)
ggcorrplot(penguins.corr, hc.order = TRUE, type = "lower", lab = TRUE) 
```
Bill Length and Flipper Length have a relatively high correlation of 0.65. This is not so high that we exclude Bill Length in a full model, 
but it is possible that it could affect the results in some way and we should keep it in mind. Now all there is to do is create our full model and test it against our reduced one. 

```{r}
gentoo.body.mass.full <- lm(body_mass_g ~ flipper_length_mm+ bill_depth_mm + bill_length_mm,data = gentoo.penguins)
summary(gentoo.body.mass.full)
anova(gentoo.body.mass, gentoo.body.mass.full)
```
These results are promising. According to our anova testing, we result in a p-value well below our significance level of 0.05, meaning we can reject the null and assume that the alternative is true: Our full model is a better fit than our reduced one. Our full model also has significant p-values for all of our variables. Our final model regarding Gentoo penguin body mass is as follows:  

$$BodyMass = -4444.42 + 23.26FlipperLength + 180.01BillDepth + 37.52BillLength$$
This means that for every 1mm increase in Flipper Length, Bill Depth, and Bill Length, we can expect approximately a 23, 180, and 38 gram increase in body mass, respectively, when holding all other variables constant. If we have a penguin with a flipper length of 220mm, bill depth of 13 mm, and bill length of 51mm, we can expect a body mass of -4444.42 + 23.26(220) + 180.01(13) + 37.52(51) = 4926.3 grams. 

```{r}
head(gentoo.penguins)

```

Suppose we wanted to compare the mass of the same species penguins across islands and sex.
Perhaps the habitat of each island could have an effect on the penguins, causing them
to be larger, or male penguins are on average larger than female penguins, for example. 
For this analysis, let's focus on only Adelie penguins,since they are the only 
species found across all three islands.  From this boxplot,we can see that there 
initially seems to be quite a different in mass between male and female penguins, 
and only a small difference between islands.
```{r}
adelie.penguins <- penguins.clean %>% filter(species == "Adelie")

adelie.penguins %>% ggplot(aes(x = island, y = body_mass_g, fill = sex)) +
  geom_boxplot() +
  xlab("Island") +
  ylab("Body Mass (g)") +
  labs(x = "Island", y = "Body Mass(g)", fill = "Sex",
       title = "Adelie penguin body mass across Biscoe, Dream, and Torgerson Islands") +
  theme_bw() 



```


Using ANOVA in R for body mass between islands confirms these results: we result in a p-value well above our significance level
of 0.05. We fail to reject the null and cannot confirm the alternative to be true.
There is not enough evidence to assume that the island has an effect on the Adelie
penguin's mass.
```{r}
adelie.islands.fit <- aov(body_mass_g ~ island, data = adelie.penguins)
summary(adelie.islands.fit)
```
Now let's see if there is a significant difference in mass for the sex of an 
Adelie penguin. We can use a simple t-test for this.  In these results, we can see
that the mean mass in grams for males is significantly larger than females.
We additionally have a p-value of nearly 0, meaning that we reject the null and
confirm the alternative as true. There is a statistically significant difference
between the mass of male and female adelie penuins.
```{r}
t.test(body_mass_g ~ sex,data = adelie.penguins)
```

## Conclusion

This was an exploratory analysis of the palmer penguins dataset where we built a linear model that could predict a penguin's body mass given its body dimensions, as well as analyzed the difference between penguin populations between islands and penguin sex. Further research of this dataset could include using logistic regression or creating a ML classifier to identify a penguin species. One thing that should ne noted about this dataset is that it is quite small. The original dataset has only a few hundred observations. The methods used in this project and especially the creation of an ML classifier would result in more accurate models if used on a larger sample size that could more accurately represent the total penguin population.

