---
title: "Penguins"
output:
  pdf_document: default
  html_document: default
date: "2025-09-26"
---

```{r setup, include=FALSE}
#tinytex::install_tinytex()
knitr::opts_chunk$set(echo = FALSE)
```
## Introduction

This is a simple analysis of the Penguins dataset from the Palmer Penguins dataset.
This analysis will go over linear regression, data visualization, anova, and t-testing. 
If you would like at the code for this project, please visit the markdown file.



## Libraries

Here are the libraries used for this viusualization and analysis.

```{r echo=TRUE, results='hide'}
library(datasets)
library(ggplot2)
library(tidyverse)
library(dplyr)
library(tibble)
library(palmerpenguins)
library(BSDA)
library(ggcorrplot)

```
## Our Dataset

This dataset contains 344 observations with 7 factors: \newline

Species: Denotes penguin species \newline
Island: Denotes an island in the Palmer Archipelago where the penguin was
observed. \newline
Bill_length_mm:  Bill length in millimeters \newline
Bill_depth_mm: Bill Depth in millimeters \newline
Flipper_length_mm: Flipper Length in millimeters \newline
Body_mass_g: Body mass in grams \newline
Sex: Denotes penguin sex (Male/Female)\newline
Year: Denotes study year(2007,2008, or 2009)

```{r}
summary(penguins)
head(penguins)
```




## Initial Visualizations

Before we dive into some regression analysis, let's get an overall understanding
of what our data looks like and how it is distributed.  With these graphs, we can
see that we have a roughly equal amount of male and female penguins. Adelie penguins
are the most plentiful, and are also the only species found on every island.

```{r}
penguins.clean <- na.omit(penguins)
penguins.clean %>% 
  ggplot(aes(x = species, fill = sex)) + 
  geom_bar(position = "dodge") +
  xlab("Island") +
  ylab("Amount") +
  labs(fill = "Sex") +
  theme_bw()

penguins.clean %>% 
  ggplot(aes(x = island, fill = species)) + 
  geom_bar(position = "dodge") +
  xlab("Island") +
  ylab("Amount") +
  labs(fill = "Species") +
  theme_bw()
  
  
```
Suppose we want to analyze if there analyze the factors behind penguin mass.
The following are graphs depicting the relationship between body dimensions and
body mass across all species of penguins. Adelie and Chinstrap Penguins
seem to have similar dimensions in regards to flipper length and bill depth.
We can also see that there seems to be some positive 
relationship between body dimenstions and body mass. 


```{r pressure, echo=FALSE}
penguins.clean %>%
  ggplot(aes(x = flipper_length_mm, y = body_mass_g, 
                        color = species, shape = species)) +
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) +
  labs(color = "Species", shape = "Species", x = "Flipper Length (mm)", 
       y = "Body Mass (g)", title = "Flipper Length vs Body Mass Across Penguin Species") +
  theme_bw()

penguins.clean %>%
  ggplot(aes(x = bill_length_mm, y = body_mass_g, 
                        color = species, shape = species)) +
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) +
   labs(color = "Species", shape = "Species", x = "Bill Length (mm)", 
       y = "Body Mass (g)", title = "Bill Length vs Body Mass Across Penguin Species") +
  theme_bw()

penguins.clean %>%
  ggplot(aes(x = bill_depth_mm, y = body_mass_g, 
                        color = species, shape = species)) +
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) +
  labs(color = "Species", shape = "Species", x = "Bill Depth (mm)", 
       y = "Body Mass (g)", title = "Bill Depth vs Body Mass Across Penguin Species") +
  labs(color = "Species", shape = "Species") +
  theme_bw()



```
Since the Gentoo seem to be isolated from the other species regarding body mass, 
let's focus on them for now. LEt's start with a simple regression analysis
using flipper length. If we do a simple regression analysis, we can see 
that flipper length is a significant predictor for body mass, with an extremely 
small p-value wellbelow our significance level of 0.05. We can reject the null and
assume the alternative is true: flipper length is a significant predictor for 
body mass. With ever 1mm increase  with flipper length for a Gentoo penguin, 
we can expect body mass to increase by 54 grams.
```{r}
gentoo.penguins <- penguins.clean %>% filter(species == "Gentoo")
gentoo.body.mass <- lm(body_mass_g ~ flipper_length_mm, data = gentoo.penguins)
summary(gentoo.body.mass)
```


Suppose we wanted to compare the mass of the same species penguins across islands.
Perhaps the habitat of each island could have an effect on the penguins, causing them
to be larger, for example. For this analysis, let's focus on only Adelie penguins,
since they are the only species found across all three islands.  From this boxplot,
we can see that there initially seems to be quite a different in mass between male
and female penguins, and only a small difference between islands.
```{r}
penguins.clean <- na.omit(penguins)
adelie.penguins <- penguins.clean %>% filter(species == "Adelie")

adelie.penguins %>% ggplot(aes(x = island, y = body_mass_g, fill = sex)) +
  geom_boxplot() +
  xlab("Island") +
  ylab("Body Mass (g)") +
  labs(x = "Island", y = "Body Mass(g)", fill = "Sex",
       title = "Adlie penguin body mass across Biscoe, Dream, and Torgerson Islands") +
  theme_bw() 



```


Using ANOVA in R confirms these results: we result in a p-value well above our p-value 
of 0.05. We fail to reject the null and cannot confirm the alternative to be true.
There is not enough evidence to assume the island has an effect on the Adelie
penguin's mass.
```{r}
adelie.islands.fit <- aov(body_mass_g ~ island, data = adelie.penguins)
summary(adelie.islands.fit)
```
Now let's see if there is a significant difference in mass for the sex of an 
Adelie penguin. We can use a simple t-test for this.  In these results, we can see
that the mean mass in grams for males is significantly larger than females.
We additionally have a p-value of nearly 0, meaning that we reject the null and
confirm the alternative as true. There is a statistically significant difference
between the mass of male and female adelie penuins.
```{r}
t.test(body_mass_g ~ sex,data = adelie.penguins)
```



